\chapter{Frequent Itemset Mining}

Frequent Itemset Mining (FIM) refers to the task of identifying itemsets in data whose frequency exceeds a given threshold. The term was first used in the context of market basket analysis in 1993~\parencite{Luna2019}. 

\section{Market Basket Analysis}

On June 26, 1974, the first grocery store Barcode scanner was used~\parencite{Chanda2019}. Barcodes soon spread to more supermarkets, enabling the systematic collection and storage of customer purchase data. This development made it possible to analyze large transactional datasets, giving rise to the market basket problem: in a large dataset of customer transactions, some items are frequently purchased together, also referred to as co-occurring or complementary items. Identifying such patterns allows supermarkets to make informed business decisions, such as optimizing product placement, promotions, and inventory~\parencite{Agrawal1993}.

\section{Definitions}

Let $I$ denote the set of all possible items (attribute values), and $D$ the set of all transactions. Using these, the following concepts can be defined:

\begin{table}[h]
\centering
\begin{tabular}{llll}
\toprule
\textbf{Concept} & \textbf{Alternative Names} & \textbf{Notation} & \textbf{Formula} \\
\midrule
Item & attribute value & $i$ & $i \in I$ \\
Itemset & \textemdash & $X$ & $X \subseteq I$ \\
Transaction & record, instance & $T$ & $T = (tid, I)$ \\
Cover & \textemdash & $\text{cover}(X)$ & $\{tid \mid (tid, I) \in D, X \subseteq I\}$ \\
Support & \textemdash & $support(X)$ & $| \text{cover}(X, D) |$ \\
Frequency & \textemdash & $P(X)$ & $\frac{support(X)}{|D|}$ \\
\bottomrule
\end{tabular}
% TODO: Citation is added as a plain text. Check if in text citation is correct and that it appears at the reference table. 
\caption{Formulas and definitions adapted from Goethals (2003).}
\label{tab:fim_definitions}
\end{table}

The itemset $X$ is a set of selected items. Each transaction $T \in D$ has a unique identifier $tid$ and its own itemset. The cover of $X$ is the set of all transaction identifiers corresponding to the transactions that contain all the items in $X$. The support of $X$ is the count of transactions that contain all the items in $X$. The frequency of $X$ is the proportion of transactions in the dataset that contain all items in $X$~\parencite{Goethals2003}.

An itemset is considered frequent if its support reaches a specified minimum threshold. This threshold can be expressed either as an absolute value $\sigma$, bounded by the size of the dataset ($0 \leq \sigma \leq |D|$), or as a relative value $\sigma_{\text{rel}} \in [0,1]$, which is scaled to the dataset size as $\sigma = \lceil \sigma_{\text{rel}} \cdot |D| \rceil$~\parencite{Goethals2003}.

\subsection{Example}
To illustrate the concepts of itemset, cover, support, and frequency, consider the small transaction dataset shown in Table.

\begin{table}[h!]
\centering
\caption{Simple transaction dataset with 6 items}
\begin{threeparttable}
\label{tab:dataset}
\begin{tabular}{|c|c|}
\hline
\textbf{TID} & \textbf{Items} \\
\hline
1 & $\{a, b\}$ \\
2 & $\{a, c\}$ \\
3 & $\{a, b, c\}$ \\
4 & $\{b\}$ \\
5 & $\{a, c, d\}$ \\
6 & $\{b, c, d\}$ \\
\hline
\end{tabular}
\begingroup
\let\item\relax
\item \textit{Source: Author.}
\endgroup
\end{threeparttable}
\end{table}

The total number of transactions in the dataset is $|D| = 6$. The following metrics are computed for the selected itemsets $\{b\}$ and $\{a, c\}$:

\begin{table}[h!]
\centering
\caption{Cover, support, and frequency for selected itemsets in the dataset}
\begin{threeparttable}
\label{tab:itemset_metrics}
\begin{tabular}{|l|c|c|}
\hline
\textbf{Metric} & \textbf{Itemset $\{b\}$} & \textbf{Itemset $\{a, c\}$} \\
\hline
Cover & $\{1, 3, 4, 6\}$ & $\{2, 3, 5\}$ \\
Support & 4 & 3 \\
Frequency & $\frac{4}{6} \approx 0.667$ & $\frac{3}{6} = 0.5$ \\
\hline
\end{tabular}
\begingroup
\let\item\relax
\item \textit{Source: Author.}
\endgroup
\end{threeparttable}
\end{table}

The following table displays all itemsets in the dataset along with their corresponding support values.

\begin{table}[h!]
\centering
\caption{Itemsets grouped by support in the dataset}
\begin{threeparttable}
\label{tab:frequency_groups}
\begin{tabular}{|c|l|}
\hline
\textbf{$support(X)$} & \textbf{Itemsets $X$} \\
\hline
4 & $\{\{a\}, \{b\}, \{c\}\}$ \\
3 & $\{\{a,c\}\}$ \\
2 & $\{\{a,b\}, \{b,c\}, \{c,d\}, \{d\}\}$ \\
1 & $\{\{a,d\}, \{b,d\}, \{a,b,c\}, \{a,c,d\}, \{b,c,d\}\}$ \\
\hline
\end{tabular}
\begingroup
\let\item\relax
\item \textit{Source: Author.}
\endgroup
\end{threeparttable}
\end{table}

Assuming a minimum support threshold of $\sigma = 3$, the set of frequent itemsets is:
\[
\{\{a\}, \{b\}, \{c\}, \{a,c\}\}.
\]

\section{Constrained Frequent Pattern Mining}

Constrained Frequent Pattern Mining considers a constraint $C$, for an itemset $I$, defined as a function
\[
C : 2^I \to \{\texttt{true}, \texttt{false}\},
\]
which is applied to itemsets. An itemset $I$ satisfies the constraint $C$ only if $C(I) = \texttt{true}$. 
The set of all itemsets that satisfy a given constraint is referred to as the \emph{theory} of the constraint, denoted as
\[
\text{Th}(C) = \{ X \in 2^I \mid C(X) = \texttt{true} \}.
\] 
\parencite{Bonchi2004}

The most typical constraints are:

\begin{itemize}
    \item \textbf{Item constraint:} Specifies particular items that should or should not be present in the itemsets.
    
    \item \textbf{Length constraint:} Specifies requirements on the number of items in the itemsets.
    
    \item \textbf{Model-based constraint:} Focuses on itemsets which are sub-itemsets or super-itemsets of given models.
    
    \item \textbf{Aggregate constraint:} Applied to an aggregate function over the items in an itemset, such as SUM, AVG, MAX, or MIN.
\end{itemize}

\parencite{Pei2002}

By definition, constraints are applied not only to itemsets themselves, but also to their subsets and individual items. It is therefore useful to characterize constraints according to several properties:

\begin{itemize}
    \item \textbf{Anti-monotone constraint:} If an itemset does not satisfy a constraint $C$, then none of its supersets satisfy $C$.
    
    \item \textbf{Monotone constraint:} If an itemset satisfies a constraint $C$, then all of its supersets also satisfy $C$.
    
    \item \textbf{Succinct constraint:} A constraint that can be evaluated by considering individual items, allowing all satisfying itemsets to be generated directly without enumerating all candidates.
    
    \item \textbf{Convertible constraint:} A constraint that can be transformed into an anti-monotone or monotone constraint under an appropriate ordering of items.
\end{itemize}

~\parencite{Pei2002}

These properties are commonly exploited to improve the efficiency of data mining algorithms, in particular by enabling effective pruning of the search space and reducing computational complexity.

https://dl.acm.org/doi/epdf/10.1145/1557019.1557030

Data Mining and Medical
Knowledge Management:
Cases and Applications

Foundations of Data Mining
and Knowledge Discovery

Charu C. Aggarwal • Jiawei Han
Editors
Frequent Pattern Mining

\section{Association Rules}

\subsection{Definition}

Assuming itemsets $X$ and $Y$, an association rule is an expression $X \Rightarrow Y$~\parencite{agrawal1996fast}.  
The confidence of this rule, sometimes referred to as accuracy, is defined as the support of the combined itemset $X \cup Y$ divided by the support of $X$ alone:
\[
\text{confidence}(X \Rightarrow Y, D) := P(Y \mid X) = \frac{\text{support}(X \cup Y, D)}{\text{support}(X, D)}
\]
~\parencite{Goethals2003}.

\subsection{Example}

Considering previous simple dataset: 

\begin{table}[h!]
\centering
\caption{Simple transaction dataset with 6 items}
\begin{threeparttable}
\label{tab:dataset}
\begin{tabular}{|c|c|}
\hline
\textbf{TID} & \textbf{Items} \\
\hline
1 & $\{a, b\}$ \\
2 & $\{a, c\}$ \\
3 & $\{a, b, c\}$ \\
4 & $\{b\}$ \\
5 & $\{a, c, d\}$ \\
6 & $\{b, c, d\}$ \\
\hline
\end{tabular}
\begingroup
\let\item\relax
\item \textit{Source: Author.}
\endgroup
\end{threeparttable}
\end{table}

\section{Apriory Alghorithm}

\section{Action Rules}